Concurrent Programming with Harmony

### 2020-07-28 book/paper.tex
Finished pass over Chapter 5
Latest

To see why, first note the invariant that if process $i$ is in the
critical section, then \textit{flags}[$i$] == \texttt{True}.
Without loss of generality,
suppose that process~0 sets \textit{flags}[0] at time $t_0$.
Process 0 can only reach the critical section if at some time $t_1$,
$t_1 > t_0$, it finds that \textit{flags}[1] == \texttt{False}.
Because of the invariant, \textit{flags}[1] == \texttt{False} implies that
process~1 is not in the critical section at time $t_1$.
Let $t_2$ be the time at which process~0 sets \textit{flags}[0]
to \texttt{False}.  Process~0 is in the critical section sometime
between $t_1$ and $t_2$.
It is easy to see that process~1 cannot enter the critical section
between $t_1$ and $t_2$, because \textit{flags}[1] == \texttt{False} at
time $t_1$.  To reach the critical section between $t_1$ and $t_2$,
it would first have to set \textit{flags}[1] to \texttt{True} and
then wait until \textit{flags}[0] == \texttt{False}.  But that does not happen
until time $t_2$.

@@

In 1981, Peterson came up with a beautiful solution to the mutual exclusion
problem, now known as ``Peterson's Algorithm''~\cite{Peterson81}.
The algorithm is an amalgam of the (incorrect) algorithms in
\autoref{fig:upflags} and \autoref{fig:upturn}, and is presented
in \autoref{fig:peterson}.
A process first indicates its interest in entering the critical section
by setting its flag.
It then politely gives way to the other process should it also want to
enter the critical section---if both do so at the same time one will
win because writes to memory in Harmony are atomic.
The process continues to be polite, waiting in a \textbf{while} loop
until either the other process is nowhere near the critical section
($\mathit{flag}[1 - \mathit{self}] == \mathtt{False}$) or has given way
($\mathit{turn} == \mathit{self}$).
Running the algorithm with Harmony shows that it satisfies both mutual
exclusion and progress.

@@

Why does it work?  We will focus here on how one might go about proving
mutual exclusion for an algorithm such as Peterson's.
For that, we have to understand a little bit more about how the Harmony
machine works.
In \autoref{ch:harmonymachine} we talked about the concept of \emph{state}:
\index{state}%
at any point in time the HVM is in a specific state.
A state is comprised of the values of the shared variables as well as
the values of the process variables
\index{process variable}%
for each process, including its
program counter and the contents of its stack.
Everytime a process executes a HVM machine instruction, the
state changes (if only because the program counter of the process
changes).  We call that a \emph{step}.
\index{step}%
Steps in Harmony are atomic.

@@

It is often useful to classify states into sets.
\emph{Initial}, \emph{final}, and \emph{reachable}, and \emph{unreachable}
are all examples of classes of states.
\autoref{fig:states} depicts a Venn diagram of various classes of states
and a trace.
One way to classify states is to define a predicate over states.
All states in which $x$ = 1, or all states where
there are two or more processes executing, are examples of such predicates.
For our purposes, it is useful to define a predicate that says that at
most one process is in the critical section.  We shall call such states
\emph{exclusive}.

@@

One way to prove that a predicate is an invariant is through induction
on the number of steps.  First you prove that the predicate holds over
the initial state.  Then you prove that for every reachable state,
and for every step from that reachable state, the predicate also holds
over the resulting state.
For this you need a predicate that describes exactly which
states are reachable.
But we do not have such a predicate: we know how to describe the set
of reachable states, but given an arbitrary state it is not easy to
see whether it is reachable or not.

To solve this problem, we will use what is called an
\emph{inductive invariant}.
\index{inductive invariant}%
An inductive invariant $\mathcal{I}$ is a predicate over states that satisfies the following:
\begin{itemize}
\item $\mathcal{I}$ holds in the initial state.
\item For any state in which $\mathcal{I}$ holds and any process in the
state takes a step, then $\mathcal{I}$ also holds in the resulting state.
\end{itemize}
Unlike an invariant, an inductive invariant must also hold over unreachable states.

One candidate for such an invariant is exclusivity itself.
After all, it certainly holds over the initial state.
And as Harmony has already determined, exclusivity is an invariant:
it holds over every reachable state.
Unfortunately, it is not an \emph{inductive} invariant.
To see why, we need to consider an \emph{unreachable} state.
It is easy to construct one: let process~0 be at label \texttt{@cs}
and process~1 at the start of the \textbf{while} loop.
Also imagine that in this particular state $\mathit{turn} = 1$.  Now let
process~1 make a sequence of steps.  Because $\mathit{turn} = 1$,
process~1 will break out of the while loop and also enter the critical
section, entering a state that is not exclusive.
So exclusivity is an invariant, but not an inductive invariant.
Doing a inductive proof with an invariant that is not inductive is usually
much harder than doing one with an invariant that is.

@@

So we are looking for an inductive invariant that \emph{implies} exclusivity.
In other words, the set of states where the inducive invariant holds
must be a subset of the set of states where there is at most one process in
the critical section.
Let $\mathcal{C}(i) = \mathit{flags}[1 - i] \land
\mathit{turn} = 1 - i$, that is, the condition on the \textbf{while} loop
for process~$i$.
Let predicate $\mathcal{I}_p(i)$ be the following.
if process $i$ is at label \texttt{@cs} (i.e., process $i$ is in the critical section),
then $\mathcal{C}(i)$ does not hold or process $1-i$ is executing after setting
$\mathit{flags}[i]$ but still before setting \textit{turn} to $1-i$.
More formally, $\mathcal{I}_p(i) = \mathtt{process}(i)@cs) \Rightarrow (\lnot \mathcal{C}(i) \lor \mathtt{process}(1-i)@gate)$.
For Peterson's Algorithm, an inductive invariant that works well is
$\mathcal{I}_p(0) \land \mathcal{I}_p(1)$.

@@

To see that $\mathcal{I}_p(0) \land \mathcal{I}_p(1)$ is, in fact, an inductive invariant, first note that
it certainly holds in the initial state, because in the initial state no process
is in the critical section.
Without loss of generality, suppose $i=0$ (a benefit from the fact that the algorithm is
symmetric for both processes).  We still have to show that if we are in a state
in which $\mathcal{I}_p(0)$ holds, then any step will result in a state in which
$\mathcal{I}_p(0)$ still holds.
If $\mathcal{I}_p(0)$ holds, process~0 is at label \texttt{@cs}.  If process~0
were to take a step, then in the next state process~0 would be no longer
at that label and $\mathcal{I}_p(0)$ would hold trivially over the next state.
Therefore we only need to consider a step by process~1.

@@

Let us consider each of these cases.
In the first case, if process~1 takes a step, there are two possibilities:
either $flags[1]$ will still be \texttt{False} (in which case the first case
continues to hold), or $flags[1]$ will be \texttt{True}
and process~1 will be at label \texttt{@gate} (in which case the third case
will hold).
We know that process~1 never sets \textit{turn} to 1, so
if the second case holds before the step, it will also hold after the step.
Finally, if process~1 is at label \texttt{@gate} before the step, then after
the step \textit{turn} will equal 0, and therefore the second case will hold
after the step.  qed.

We have now demonstrated mutual exclusion in Peterson's Algorithm in two
different ways: one by letting Harmony explore all possible executions, the
other using an inductive invariant and proof by induction.  The former
is certainly easier, but it does not provide intuition for why the
algorithm works.  The second provides much more insight.  We therefore
encourage to include inductive invariants in your Harmony code.

@@

This anecdote suggests the following.  If you need to do an induction
proof of an algorithm, you have to come up with an inductive invariant.
Before trying to prove the algorithm, you can check that the predicate is
at least invariant by testing it using Harmony.  Doing so could potentially
avoid wasting your time on a proof that will not work because the
predicate is not invariant, and therefore not an inductive invariant either.
(The author fixed the Wikipedia page.)

>>>>

To see why, first note the following invariant: if process $i$ is in the
critical section, then \textit{flags}[$i$] == \texttt{True}.
Without loss of generality,
suppose that process~0 sets \textit{flags}[0] at time $t_0$.
Process 0 can only reach the critical section if at some time $t_1$,
$t_1 > t_0$, it finds that \textit{flags}[1] == \texttt{False}.
Because of the invariant, \textit{flags}[1] == \texttt{False} implies that
process~1 is not in the critical section at time $t_1$.
Let $t_2$ be the time at which process~0 sets \textit{flags}[0]
to \texttt{False}.  Process~0 is in the critical section sometime
between $t_1$ and $t_2$.
It is easy to see that process~1 cannot enter the critical section
between $t_1$ and $t_2$, because \textit{flags}[1] == \texttt{False} at
time $t_1$.  To reach the critical section between $t_1$ and $t_2$,
it would first have to set \textit{flags}[1] to \texttt{True} and
then wait until \textit{flags}[0] == \texttt{False}.  But that does not happen
until time $t_2$.

@@

In 1981, Gary L.~Peterson came up with a beautiful solution to the mutual exclusion
problem, now known as ``Peterson's Algorithm''~\cite{Peterson81}.
The algorithm is an amalgam of the (incorrect) algorithms in
\autoref{fig:upflags} and \autoref{fig:upturn}, and is presented
in \autoref{fig:peterson}.
A process first indicates its interest in entering the critical section
by setting its flag.
It then politely gives way to the other process should it also want to
enter the critical section---if both do so at the same time one will
win because writes to memory in Harmony are atomic.
The process continues to be polite, waiting in a \textbf{while} loop
until either the other process is nowhere near the critical section
($\mathit{flag}[1 - \mathit{self}] == \mathtt{False}$) or has given way
($\mathit{turn} == \mathit{self}$).
Running the algorithm with Harmony shows that it satisfies both mutual
exclusion and progress.

@@

Why does it work?  We will focus here on how one might go about proving
mutual exclusion for an algorithm such as Peterson's.
For that, we have to understand a little bit more about how the Harmony
virtual machine (HVM) works.
In \autoref{ch:harmonymachine} we talked about the concept of \emph{state}:
\index{state}%
at any point in time the HVM is in a specific state.
A state is comprised of the values of the shared variables as well as
the values of the process variables
\index{process variable}%
of each process, including its
program counter and the contents of its stack.
Each time a process executes a HVM machine instruction, the
state changes (if only because the program counter of the process
changes).  We call that a \emph{step}.
\index{step}%
Steps in Harmony are atomic.

@@

It is often useful to classify states.
\emph{Initial}, \emph{final}, and \emph{reachable}, and \emph{unreachable}
are all examples of classes of states.
\autoref{fig:states} depicts a Venn diagram of various classes of states
and a trace.
One way to classify states is to define a predicate over states.
All states in which $x$ = 1, or all states where
there are two or more processes executing, are examples of such predicates.
For our purposes, it is useful to define a predicate that says that at
most one process is in the critical section.  We shall call such states
\emph{exclusive}.

@@

One way to prove that a predicate is an invariant is through induction
on the number of steps.  First you prove that the predicate holds over
the initial state.  Then you prove that for every reachable state,
and for every step from that reachable state, the predicate also holds
over the resulting state.
For this to work you would need a predicate that describes exactly which
states are reachable.
But we do not have such a predicate: we know how to describe the set
of reachable states, but given an arbitrary state it is not easy to
see whether it is reachable or not.

To solve this problem, we will use what is called an
\emph{inductive invariant}.
\index{inductive invariant}%
An inductive invariant $\mathcal{I}$ is a predicate over states that satisfies the following:
\begin{itemize}
\item $\mathcal{I}$ holds in the initial state.
\item For any state in which $\mathcal{I}$ holds and for any process in the
state, if the process takes a step, then $\mathcal{I}$ also holds in the resulting state.
\end{itemize}
Unlike an ordinary invariant, an inductive invariant must also hold over unreachable states.

One candidate for such an invariant is exclusivity itself.
After all, it certainly holds over the initial state.
And as Harmony has already determined, exclusivity is an invariant:
it holds over every reachable state.
Unfortunately, it is not an \emph{inductive} invariant.
To see why, we need to consider an \emph{unreachable} state.
It is easy to construct one: let process~0 be at label \texttt{@cs}
and process~1 be at the start of the \textbf{while} loop.
Also imagine that in this particular state $\mathit{turn} = 1$.  Now let
process~1 make a sequence of steps.  Because $\mathit{turn} = 1$,
process~1 will break out of the while loop and also enter the critical
section, entering a state that is not exclusive.
So exclusivity is an invariant, but not an inductive invariant.
Doing a inductive proof with an invariant that is not inductive is usually
much harder than doing one with an invariant that is.

@@

So we are looking for an inductive invariant that \emph{implies} exclusivity.
In other words, the set of states where the inducive invariant holds
must be a subset of the set of states where there is at most one process in
the critical section.
Let $\mathcal{C}(i) = \mathit{flags}[1 - i] \land
\mathit{turn} = 1 - i$, that is, the condition on the \textbf{while} loop
for process~$i$.
Let predicate $\mathcal{I}_p(i)$ be the following.
if process $i$ is at label \texttt{@cs} (i.e., process $i$ is in the critical section),
then $\mathcal{C}(i)$ does not hold or process $1-i$ is executing after setting
$\mathit{flags}[i]$ but still before setting \textit{turn} to $1-i$.
More formally, $\mathcal{I}_p(i) = \mathtt{process}(i)@cs \Rightarrow (\lnot \mathcal{C}(i) \lor \mathtt{process}(1-i)@gate)$.
For Peterson's Algorithm, an inductive invariant that works well is
$\mathcal{I}_p(0) \land \mathcal{I}_p(1)$.

@@

To see that $\mathcal{I}_p(0) \land \mathcal{I}_p(1)$ is, in fact, an inductive invariant, first note that
it holds in the initial state, because in the initial state no process
is in the critical section.
Without loss of generality, suppose $i=0$ (a benefit from the fact that the algorithm is
symmetric for both processes).  We still have to show that if we are in a state
in which $\mathcal{I}_p(0)$ holds, then any step will result in a state in which
$\mathcal{I}_p(0)$ still holds.
If $\mathcal{I}_p(0)$ holds, process~0 is at label \texttt{@cs}.  If process~0
were to take a step, then in the next state process~0 would be no longer
at that label and $\mathcal{I}_p(0)$ would hold trivially over the next state.
Therefore we only need to consider a step by process~1.

@@

Let us consider each of these cases.
We have to show that if process~1 takes a step, then one of those
cases must hold after the step.
In the first case, if process~1 takes a step, there are two possibilities:
either $flags[1]$ will still be \texttt{False} (in which case the first case
continues to hold), or $flags[1]$ will be \texttt{True}
and process~1 will be at label \texttt{@gate} (in which case the third case
will hold).
We know that process~1 never sets \textit{turn} to 1, so
if the second case holds before the step, it will also hold after the step.
Finally, if process~1 is at label \texttt{@gate} before the step, then after
the step \textit{turn} will equal 0, and therefore the second case will hold
after the step.  qed.

We have now demonstrated mutual exclusion in Peterson's Algorithm in two
different ways: one by letting Harmony explore all possible executions, the
other using an inductive invariant and proof by induction.  The former
is certainly easier, but it does not provide intuition for why the
algorithm works.  The second provides much more insight.  We therefore
encourage to include invariants, particularly inductive ones, in your
Harmony code.

@@

This anecdote suggests the following.  If you need to do a proof by induction
of an algorithm, you have to come up with an inductive invariant.
Before trying to prove the algorithm, you can check that the predicate is
at least invariant by testing it using Harmony.  Doing so could potentially
avoid wasting your time on a proof that will not work because the
predicate is not invariant, and therefore not an inductive invariant either.
(The author fixed the Wikipedia page with the help of Fred B.~Schneider.)
